---
title: "DissociationTest.Rmd"
author: "Rayna M Harris"
date: November 20, 2017
#output: html_document
output:
  md_document:
    variant: markdown_github
---

This R Markdown document will walk through the analysis of hippocampal tissue prepared with two different methods. The "homogenized" samples were collected by punch then homogenized in homogenization buffer from the Promega Maxwell kit. The "dissociated samples" were also collected similarily but the cells was dissociated after being punch and before being homogenized. 

```{r setup, echo=FALSE, message=FALSE, results='hide',comment=FALSE, warning=FALSE}
#source("http://www.bioconductor.org/biocLite.R")
#biocLite("DESeq2")
library(DESeq2)
library(magrittr)
library(tidyverse)
library(reshape2)
library(VennDiagram)
library(genefilter)
library(pheatmap)
library(colorRamps) # for a matlab like color scheme

# set output file for figures 
knitr::opts_chunk$set(fig.path = '../figures/CA1_allgroups/')
```

```{r ImportData, echo=FALSE, message=FALSE}
# this starts with data genearated from code described in KallistoGather.Rmd
colData <- read.csv('../intermediatefiles/colData.csv')
countData <-  read.csv('../intermediatefiles/countData.csv', check.names = F)

```

```{r SubsetData, echo=FALSE, message=FALSE}
#### now slim the data to get some new figures 
colData <- colData %>% 
  #filter(grepl("Homogenized|Dissociated", Method))  %>% 
  filter(Punch == "CA1")  %>% 
  droplevels() ## subsets data
savecols <- as.character(colData$RNAseqID) #select the sample name column that corresponds to row names
savecols <- as.vector(savecols) # make it a vector
countData <- countData %>% select(one_of(savecols)) # select just the columns that match the samples in colData
```



#### Differential Gene Expression Plots
```{r DifferentialGeneExpressionAnalysis, echo=FALSE, message=FALSE, warning=FALSE}
dds <- DESeqDataSetFromMatrix(countData = countData,
                              colData = colData,
                              design = ~ Method )
## filter genes with 0 counts
dds <- dds[ rowSums(counts(dds)) > 1, ]
dds

# Differential expression analysis
dds <- DESeq(dds)
dds

# general deseq
res <- results(dds, independentFiltering = F)
#res
summary(res)
resOrdered <- res[order(res$padj),]
#sum(res$padj < 0.1, na.rm = TRUE)
res05 <- results(dds, alpha=0.05)
summary(res05)
#sum(res05$padj < 0.05, na.rm=TRUE)

## make a plot of fold change as function of expression

plot <- plotMA(res05, main="MA Plot")
return(plot)

resMLE <- results(dds)
#head(resMLE, 4)

## plot most significatnly expressed gene
plotCounts(dds, gene=which.min(res$padj), intgroup="Punch")
# panel2: STXBP6 binds components of the SNARE complex
# panel1: crlf1 Cytokine Receptor Like Factor 1 
plotCounts(dds, gene=which.min(res$padj), intgroup="Method")
# CA1 homodiss only selplg 
# acta DG all groups

## for variance stablized gene expression and log transformed data
rld <- rlog(dds, blind=FALSE)
vsd <- varianceStabilizingTransformation(dds, blind=FALSE)
vsd.fast <- vst(dds, blind=FALSE)

## DEG by contrasts
resMethodHomogDiss <- results(dds, contrast = c("Method", "Homogenized", "Dissociated"), independentFiltering = F)
#sum(resMethodHomogDiss$padj < 0.1, na.rm = TRUE) #304 #554 #435
valsMethodHomogDiss <- cbind(resMethodHomogDiss$pvalue, resMethodHomogDiss$padj)
colnames(valsMethodHomogDiss)=c("pval.MethodHomogDiss", "padj.MethodHomogDiss")

resMethodYokedTrained <- results(dds, contrast = c("Method", "Yoked", "Trained"), independentFiltering = F)
#sum(resMethodYokedTrained$padj < 0.1, na.rm = TRUE) #4
valsMethodYokedTrained <- cbind(resMethodYokedTrained$pvalue, resMethodYokedTrained$padj) 
colnames(valsMethodYokedTrained)=c("pval.MethodYokedTrained", "padj.MethodYokedTrained")

#head(assay(rld), 3)
rldd <- assay(rld)
rldpvals <- cbind(rldd, valsMethodHomogDiss, valsMethodYokedTrained)

```


resPunchCA1DG <- results(dds, contrast = c("Punch", "CA1", "DG"), independentFiltering = F)
#sum(resPunchCA1DG$padj < 0.1, na.rm = TRUE) # 4170 #1127
valsPunchCA1DG <- cbind(resPunchCA1DG$pvalue, resPunchCA1DG$padj) 
colnames(valsPunchCA1DG)=c("pval.CA1DG", "padj.CA1DG")

resPunchCA1CA3 <- results(dds, contrast = c("Punch", "CA1", "CA3"), independentFiltering = F)
#sum(resPunchCA1CA3$padj < 0.1, na.rm = TRUE) #2240 # 70
valsPunchCA1CA3 <- cbind(resPunchCA1CA3$pvalue, resPunchCA1CA3$padj) 
colnames(valsPunchCA1CA3)=c("pval.CA1CA3", "padj.CA1CA3")

resPunchCA3DG <- results(dds, contrast = c("Punch", "CA3", "DG"), independentFiltering = F)
#sum(resPunchCA3DG$padj < 0.1, na.rm = TRUE) #4785 #591
valsPunchCA3DG <- cbind(resPunchCA3DG$pvalue, resPunchCA3DG$padj) 
colnames(valsPunchCA3DG)=c("pval.CA3DG", "padj.CA3DG")


#```{r VennDiagram, echo=FALSE, message=FALSE}
rldpvals <- as.data.frame(rldpvals)

MethodHomogDiss <- row.names(rldpvals[rldpvals$padj.MethodHomogDiss<0.1 & !is.na(rldpvals$padj.MethodHomogDiss),])
#MethodYokedTrained <- row.names(rldpvals[rldpvals$padj.valsMethodYokedTrained<0.1 & !is.na(rldpvals$padj.valsMethodYokedTrained),])
PunchCA1DG <- row.names(rldpvals[rldpvals$padj.CA1DG<0.1 & !is.na(rldpvals$padj.CA1DG),])
PunchCA1CA3 <- row.names(rldpvals[rldpvals$padj.CA1CA3<0.1 & !is.na(rldpvals$padj.CA1CA3),])
PunchCA3DG <- row.names(rldpvals[rldpvals$padj.CA3DG<0.1 & !is.na(rldpvals$padj.CA3DG),])

## four way grid
candidates <- list("CA1 v. DG" = PunchCA1DG, "CA1 v. CA3" = PunchCA1CA3, "CA3 v. DG" = PunchCA3DG,  "Homogenized v. Dissociated" = MethodHomogDiss )
dev.off()
prettyvenn <- venn.diagram(
  x = candidates, filename=NULL, lwd=4,
  col = "transparent",
  fill = (values=c("#00441b", "#00441b","#238b45", "#238b45")),
  alpha = 0.5,
  cex = 1, fontfamily = "sans", #fontface = "bold",
  cat.default.pos = "text",
  #cat.col = c("darkred", "darkgreen", "blue4", "orange"),
  #cat.dist = c(0.08, 0.08, 0.08, 0.08), cat.pos = 1,
  cat.cex = 1, cat.fontfamily = "sans")
grid.draw(prettyvenn)

## ca1 ca3 homo diss

candidates <- list("CA1 v. CA3" = PunchCA1CA3,  "Homogenized v. Dissociated" = MethodHomogDiss )
dev.off()
prettyvenn <- venn.diagram(
  x = candidates, filename=NULL, lwd=2,
  col = "transparent",
  fill = (values=c("#00441b", "#00441b")),
  alpha = 0.5,
  cex = 1, fontfamily = "sans", #fontface = "bold",
  cat.default.pos = "text",
  #cat.col = c("darkred", "darkgreen", "blue4", "orange"),
  #cat.dist = c(0.08, 0.08, 0.08, 0.08), cat.pos = 1,
  cat.cex = 1, cat.fontfamily = "sans")
grid.draw(prettyvenn)

## ca1 dg homo diss
candidates <- list("CA1 v. DG" = PunchCA1DG,  "Homogenized v. Dissociated" = MethodHomogDiss )
dev.off()
prettyvenn <- venn.diagram(
  x = candidates, filename=NULL, lwd=4,
  col = "transparent",
  fill = (values=c("#00441b", "#00441b")),
  alpha = 0.5,
  cex = 1, fontfamily = "sans", #fontface = "bold",
  cat.default.pos = "text",
  #cat.col = c("darkred", "darkgreen", "blue4", "orange"),
  #cat.dist = c(0.08, 0.08, 0.08, 0.08), cat.pos = 1,
  cat.cex = 1, cat.fontfamily = "sans")
grid.draw(prettyvenn)

## ca3 dg homo diss
candidates <- list("CA3 v. DG" = PunchCA3DG,  "Homogenized v. Dissociated" = MethodHomogDiss )
dev.off()
prettyvenn <- venn.diagram(
  x = candidates, filename=NULL, lwd=2,
  col = "transparent",
  fill = (values=c("#00441b", "#00441b")),
  alpha = 0.5,
  cex = 1, fontfamily = "sans", #fontface = "bold",
  cat.default.pos = "text",
  #cat.col = c("darkred", "darkgreen", "blue4", "orange"),
  #cat.dist = c(0.08, 0.08, 0.08, 0.08), cat.pos = 1,
  cat.cex = 1, cat.fontfamily = "sans")
grid.draw(prettyvenn)

#```

```{r Heatmap100DEgenes, echo=FALSE, message=FALSE}
nt <- normTransform(dds) # defaults to log2(x+1) 
df <- as.data.frame(colData(dds)[,c("Method", "Punch")])
ann_colors = list(Method = c(Homogenized = (values=c("#969696")), Dissociated = (values=c("#525252")), Trained = (values=c("#542788")), Yoked = (values=c("#f1a340"))),
  Punch =  c(DG = (values=c("#980043")),  CA3 = (values=c("#c994c7")), 
             CA1 = (values=c("#dd1c77"))))
matlabcolors <-  matlab.like2(100)  #color schele

#topVarGenes <- head(order(rowVars(assay(rld)),decreasing=TRUE),70)
#mat <- assay(rld)[ topVarGenes, ]
#mat <- mat - rowMeans(mat)

DEGes <- as.data.frame(rldpvals) # convert matrix to dataframe
DEGes$rownames <- rownames(DEGes)  # add the rownames to the dataframe
DEGes$padjmin <- with(DEGes, pmin(padj.MethodHomogDiss, padj.MethodYokedTrained)) # put the min pvalue in a new column
DEGes <- DEGes %>% filter(padj.MethodYokedTrained < 0.05)
rownames(DEGes) <- DEGes$rownames
drop.cols <- c("padj.MethodHomogDiss", "pval.MethodHomogDiss","padj.MethodYokedTrained", "pval.MethodYokedTrained", "rownames", "padjmin")
DEGes <- DEGes %>% select(-one_of(drop.cols))
DEGes <- as.matrix(DEGes)
DEGes <- DEGes - rowMeans(DEGes)

pheatmap(DEGes, show_colnames=F, show_rownames = TRUE,
         annotation_col=df, annotation_colors = ann_colors,
         fontsize = 12, fontsize_row = 10, 
         #cellwidth=10, cellheight=10, width = 10,
         border_color = "grey60" ,
         color = matlabcolors,
         main = "Trained vs Yoked padj < 0.05"
         )

DEGes <- as.data.frame(rldpvals) # convert matrix to dataframe
DEGes$rownames <- rownames(DEGes)  # add the rownames to the dataframe
DEGes$padjmin <- with(DEGes, pmin(padj.MethodHomogDiss, padj.MethodYokedTrained)) # put the min pvalue in a new column
DEGes <- DEGes %>% filter(padj.MethodHomogDiss < 0.05)
rownames(DEGes) <- DEGes$rownames
drop.cols <- c("padj.MethodHomogDiss", "pval.MethodHomogDiss","padj.MethodYokedTrained", "pval.MethodYokedTrained", "rownames", "padjmin")
DEGes <- DEGes %>% select(-one_of(drop.cols))
DEGes <- as.matrix(DEGes)
DEGes <- DEGes - rowMeans(DEGes)

pheatmap(DEGes, show_colnames=F, show_rownames = TRUE,
         annotation_col=df, annotation_colors = ann_colors,
         fontsize = 12, fontsize_row = 10, 
         #cellwidth=10, cellheight=10, width = 10,
         border_color = "grey60" ,
         color = matlabcolors,
         main = "Homogenized vs Dissociated padj < 0.05"
         )

DEGes <- as.data.frame(rldpvals) # convert matrix to dataframe
DEGes$rownames <- rownames(DEGes)  # add the rownames to the dataframe
DEGes$padjmin <- with(DEGes, pmin(padj.MethodHomogDiss, padj.MethodYokedTrained)) # put the min pvalue in a new column
DEGes <- DEGes %>% filter(padjmin < 0.05)
rownames(DEGes) <- DEGes$rownames
drop.cols <- c("padj.MethodHomogDiss", "pval.MethodHomogDiss","padj.MethodYokedTrained", "pval.MethodYokedTrained", "rownames", "padjmin")
DEGes <- DEGes %>% select(-one_of(drop.cols))
DEGes <- as.matrix(DEGes)
DEGes <- DEGes - rowMeans(DEGes)

pheatmap(DEGes, show_colnames=F, show_rownames = TRUE,
         annotation_col=df, annotation_colors = ann_colors,
         fontsize = 12, fontsize_row = 10, 
         #cellwidth=10, cellheight=10, width = 10,
         border_color = "grey60" ,
         color = matlabcolors,
         main = "Any padj < 0.05"
         )


```

```{r PCA, echo=FALSE, message=FALSE}
source("~/Github/BehavEphyRNAseq/bin/02f_DESeqPCAfunction.R")
plotPCA(rld, intgroup=c("Method", "Punch"), returnData=TRUE)
pcadata <- plotPCA(rld, intgroup=c("Method", "Punch"), returnData=TRUE)
percentVar <- round(100 * attr(pcadata, "percentVar"))

ggplot(pcadata, aes(PC1, PC2, color=Punch, shape=Method)) +
  geom_point(size=5) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) 

```