---
title: "volcanos"
author: "Rayna M Harris"
date: "12/16/2017"
output: md_document
---

These are the packages I need for my volcano plots.

```{r setup, include=FALSE}
library(ggplot2)
library(cowplot)

# set output file for figures 
knitr::opts_chunk$set(fig.path = '../figures/volcanoplots/')
```

Here I load the relevant dataframes and set the color palette.

```{r }
cembrowski <- read.csv("../results/04_cembrowski_volcanoTreatment.csv", header = T, row.names = 1)
cembrowski$wrap <- "Ventral <-> Dorsal"
cembrowskicolor <- c("ventral" = "#525252", "dorsal" = "#525252", "none" = "#d9d9d9")


cognition <- read.csv("../results/03_cognition_volcanoTreatment.csv", header = T, row.names = 1)
cognition$wrap <- "yoked <-> trained"
cognitioncolor <-  c("trained" = "#525252", "yoked" = "#525252", "none" = "#d9d9d9")

stress <- read.csv("../results/02_stress_volcanoTreatment.csv", header = T, row.names = 1)
stress$wrap <- "homecage <-> stressed"
stresscolor <-  c("shocked" = "#525252", "homecage" = "#525252", "none" = "#d9d9d9")

dissociation <- read.csv("../results/01_dissociation_volcanoTreatment.csv", header = T, row.names = 1)
dissociation$wrap <- "homogenized <-> dissociated"
dissociationcolor <-  c("dissociated" = "#525252", "control" = "#525252", "none" = "#d9d9d9")

```

Here's my function for plotting the effect of treatment. Since each treatment has different levels, I set the color code outside the function.

```{r }
VolcanoTreatment <- function(filename, colorval){
  data <- filename
  volcanoplot <- ggplot(data, aes(x = lfc, y = pvalue)) + 
  geom_point(aes(color = factor(color), shape = factor(color)), 
             size = 1, alpha = 0.8, na.rm = T) +
  theme_bw(base_size = 8) + # clean up theme
  theme(legend.position = "none",
        panel.grid.minor=element_blank(),
        panel.grid.major=element_blank())+
  scale_color_manual(values = colorval) +
  scale_x_continuous(name="log fold change",
                      limits = c(-10, 10)) +
  scale_y_continuous(name="-log10 p-value",
                     limits = c(0, 6),
                     breaks = c(1,6)) +
  geom_hline(yintercept = 1,  size = 0.25, linetype = 2 ) + 
  scale_shape_manual(values = c(1,16,16)) +
    facet_wrap(~wrap)
  plot(volcanoplot)
  myfile = paste("../figures/volcanoplots/Treatment_", substitute(filename), ".pdf", sep="")
  pdf(file = myfile, width=1.75, height=2)
  plot(volcanoplot)
  dev.off()
}
```

Now, let's run the function on the four files.

```{r Treatment}
VolcanoTreatment(cembrowski, cembrowskicolor)
VolcanoTreatment(cognition, cognitioncolor)
VolcanoTreatment(stress, stresscolor)
VolcanoTreatment(dissociation, dissociationcolor)
```

Now this is the fucntion for plotting CA1 vs. DG volcano plots. The color here is set inside. 

```{r }
VolcanoCA1DG <- function(filename){
  data <- filename
  volcanoplot <- ggplot(data, aes(x = lfc, y = pvalue)) + 
  geom_point(aes(color = factor(color), shape = factor(color)), 
             size = 1, alpha = 0.8, na.rm = T) + # add gene points
  theme_bw(base_size = 8) + # clean up theme
  theme(legend.position = "none", # remove legend 
        panel.grid.minor=element_blank(),
        panel.grid.major=element_blank()) + 
  scale_color_manual(values = c("CA1" = "#7570b3",
                                "DG" = "#d95f02", 
                                "none" = "#d9d9d9")) +   
  scale_x_continuous(name="log fold change",
                     limits = c(-10, 10)) +
  scale_y_continuous(name="-log10 p-value",
                     limits = c(0, 18),
                     breaks = c(1,6,12,18)) +
  geom_hline(yintercept = 1,  size = 0.25, linetype = 2 ) + 
  scale_shape_manual(values = c(16,16,16)) +
    facet_wrap(~wrap)
  plot(volcanoplot)
  myfile = paste("../figures/volcanoplots/CA1DG_", substitute(filename), ".pdf", sep="")
  pdf(file = myfile, width=1.75, height=2)
  plot(volcanoplot)
  dev.off()
}
VolcanoCA1DG(dissociation)

```

The files

```{r }
cembrowski <- read.csv("../results/04_cembrowski_volcanoCA1DG.csv", header = T, row.names = 1)
cembrowski$wrap <- "Dorsal Ventral Gradient"

cognition <- read.csv("../results/03_cognition_volcanoCA1DG.csv", header = T, row.names = 1)
cognition$wrap <- "Cognitive Training"

stress <- read.csv("../results/02_stress_volcanoCA1DG.csv", header = T, row.names = 1)
stress$wrap <- "Stressful Enviornment"

dissociation <- read.csv("../results/01_dissociation_volcanoCA1DG.csv", header = T, row.names = 1)
dissociation$wrap <- "DG <-> CA1"
```

```{r CA1DG}
VolcanoCA1DG(cembrowski)
VolcanoCA1DG(cognition)
VolcanoCA1DG(stress)
VolcanoCA1DG(dissociation)
```

## Useful R tutorials
- http://ggplot2.tidyverse.org/reference/scale_continuous.html for adjusting ggplot axis